---
title: "import-quantification-data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{import-quantification-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=FALSE
)
```

# Importing Quantification Data

## Introduction

This vignette describes how to take the RSEM quantification, files which are 
output by the nf-core/rnaseq pipeline, and collect them into a `DeseqDataSet` 
object. The `DeseqDataSet` object inherits from the Bioconductor base class 
`SummarizedExperiment`, and so is generally a useful container into which to 
put this data, even if you do not plan to use `DESeq`.  

Note that after creating the samplesheet, you should use it to run the pipeline, 
but also to update the RNASeq database with the new samples.

See the script
[R/import_quantification_data.R](../R/import_quantification_data.R), which
includes an example of using the package rslurm to easily run the data
aggregation on a cluster with SLURM.

## combining txi objects

Combining txi objects (eg, to aggregate a batch when it comes in and combine 
it with the larger set), see
[R/combine_txi_objects.R](../R/combine_txi_objects.R)

## Creating a DDS object

```{r create_dds}

library(tidyverse)
library(DESeq2)
library(rtracklayer)
library(here)

# set variables

WRITE = TRUE

FORMULA = ~batch+sex

# whether to create the gene or tx level object. set to TRUE for tx
TX = TRUE

# load data --------------------------------------------------------------------

con = RSQLite::dbConnect(RSQLite::SQLite(),here("llfs_rnaseq_data/llfs_rnaseq_database.sqlite"))

txi_list = list(
  gene = here('/scratch/mblab/chasem/data/txi_combined_gene_20231023_final.rds'),
  tx = here('/scratch/mblab/chasem/data/txi_combined_tx_20231023_final.rds')
)

if(TX){
  txi = readRDS(txi_list$tx)
  DDS_LEGAL_OUTPUT = "/scratch/mblab/chasem/data/dds_tx_legal_20231023_final.rds"

} else{
  txi = readRDS(txi_list$gene)
  DDS_LEGAL_OUTPUT = "/scratch/mblab/chasem/data/dds_gene_legal_20231023_final.rds"
}

meta_df = tbl(con,'llfs_metadata_subject') %>%
  mutate(subject = as.character(subject),
         visitcode=as.character(visitcode)) %>%
  collect() %>%
  mutate(count_headers = paste0('library_', library_id),
         sex = as.character(sex)) %>%
  mutate(batch = factor(batch),
         sex = factor(sex))

setdiff1 = length(setdiff(meta_df$count_headers, colnames(txi$abundance)))
setdiff2 = length(setdiff(colnames(txi$abundance), meta_df$count_headers))

stopifnot(setdiff1 == setdiff2 & setdiff1 == 0)

txi$abundance = txi$abundance[,meta_df$count_headers]
txi$counts = txi$counts[,meta_df$count_headers]
txi$length = txi$length[,meta_df$count_headers]

stopifnot(identical(meta_df$count_headers, colnames(txi$abundance)))

# add row ranges
gencode_38 = readRDS("/ref/mblab/data/human/GRCh38/gencode/v38/gencode.v38.annotation_granges.rds") 
# rtracklayer::import("/ref/mblab/llfs_rnaseq_data/human/GRCh38/gencode/v38/gencode.v38.annotation.gtf")

# filter the annotations depending on whether gene or tx quants
if(TX){
  gencode_38_fltr = gencode_38$transcript_id %in% rownames(txi$abundance)
  ranges = gencode_38[gencode_38_fltr]
  ranges = ranges[ranges$type == "transcript"]

  # reorder and check
  ranges = ranges[order(match(ranges$transcript_id,rownames(txi$abundance)))]
  stopifnot(identical(ranges$transcript_id, rownames(txi$abundance)))
} else{
  gencode_38_fltr = gencode_38$gene_id %in% rownames(txi$abundance)
  ranges = gencode_38[gencode_38_fltr]
  ranges = ranges[ranges$type == "gene"]

  # reorder and check
  ranges = ranges[order(match(ranges$gene_id,rownames(txi$abundance)))]
  stopifnot(identical(ranges$gene_id, rownames(txi$abundance)))
}

## unfiltered DDS object ----
dds = DESeqDataSetFromTximport(txi = txi,
                               colData = meta_df,
                               rowRanges = ranges,
                               design = FORMULA)

rownames(dds) = rownames(txi$abundance)
colnames(dds) = dds$count_headers

# remove samples for which we do not have legal permission
dds_legal = dds[,dds$legal == 1]


if(WRITE){
  #write_rds(dds,DDS_FULL_OUTPUT)
  write_rds(dds_legal,DDS_LEGAL_OUTPUT)
}


```

## Misc

### Create tx2gene

The transcript to gene file is just as it sounds -- a map from transcript
identifiers to gene identifiers. This is used by tximport.

```{r tx2gene}
library(tidyverse)

# this is v38 from gencode
gtf = read_tsv("llfs_rnaseq_data/gencode.v38.annotation.gtf", comment="#", col_names = FALSE)

btwn_quotes_regex = '"(.*?)"'
remove_prefix_and_quotes_regex = '\ |\"'

gtf_parsed = gtf %>%
  # separate(X9, into=c("gene_id", "transcript_id", "other"), extra="merge", sep=";") %>%
  mutate(gene_id = str_extract(X9, paste0("gene_id\ ", btwn_quotes_regex)),
         transcript_id = str_extract(X9, paste0("transcript_id\ ", btwn_quotes_regex)),
         gene_name = str_extract(X9, paste0("gene_name\ ", btwn_quotes_regex)),
         gene_type = str_extract(X9, paste0("gene_type\ ", btwn_quotes_regex)),
         exon_id = str_extract(X9, paste0("exon_id\ ", btwn_quotes_regex)),
         exon_number = str_extract(X9, paste0("exon_number\ ", btwn_quotes_regex))) %>%
  dplyr::rename(chromosome = X1,
                source = X2,
                sequence_ontology = X3,
                start = X4,
                stop = X5,
                strand = X7,
                phase = X8,
                description_string = X9) %>%
  mutate(gene_id = str_remove_all(gene_id, paste0('gene_id', remove_prefix_and_quotes_regex)),
         transcript_id = str_remove_all(transcript_id, paste0('transcript_id', remove_prefix_and_quotes_regex)),
         gene_name = str_remove_all(gene_name, paste0('gene_name', remove_prefix_and_quotes_regex)),
         gene_type = str_remove_all(gene_type, paste0('gbkey', remove_prefix_and_quotes_regex)),
         exon_id = str_remove_all(exon_number, paste0('exon_id', remove_prefix_and_quotes_regex)),
         exon_number = str_remove_all(exon_number, paste0('exon_number', remove_prefix_and_quotes_regex))) %>%
  dplyr::select(chromosome,
                source,
                sequence_ontology,
                start,
                stop,
                X6,
                strand,
                phase,
                gene_id,
                transcript_id,
                gene_name,
                gene_type,
                exon_id,
                exon_number)

# inspection showed that ___ duplicated tx go to the same gene id
tx2gene = gtf_parsed %>%
  filter(sequence_ontology == "exon") %>%
  dplyr::select(transcript_id, gene_id, gene_type, gene_name, exon_id, exon_number) %>%
  distinct(transcript_id, .keep_all = TRUE)

# write_rds(gtf_parsed, "llfs_rnaseq_data/gencode38_gtf_parsed_as_df_20210919.csv")
# 
# write_csv(tx2gene, "llfs_rnaseq_data/gencode38_tx2gene_20210919.csv")

```

