---
title: "sex-label-check"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sex-label-check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=FALSE,
  dpi=200,
  fig.width = 14,
  fig.height = 8
)

```

# Check for sex mislabels

Sex mislabeling implies sample mislabels, so here we're looking for 
samples which group with the opposite of their sex label. Sorry libs, this is 
quite binary.

It happens to be the case that using the top two PCs calculated on the top 
500 most variable genes very cleanly separates the samples by sex. We use this 
to do this analysis below:

```{r, echo=FALSE}
library(tidyverse)
library(ggrepel)
library(DESeq2)
library(edgeR)
library(caret)
library(ggsci)
library(RSQLite)
library(here)
library(llfsRnaseq)

# set ggplot global options
theme_set(theme_minimal())
theme_update(text = element_text(size=20),
             panel.border = element_rect(colour = "black", fill=NA, linewidth=1))


DDS_RDS_PATH = "/scratch/mblab/chasem/data/dds_gene_legal_20231023_final.rds"

# remove data objects which are not used after a given chunk
CLEAN_UP = TRUE

OVERWRITE_FLAG = FALSE

WRITE_OUT = FALSE


qc_datasets = sex_mislabel_setup(DDS_RDS_PATH, .015, 3)


slope=3.5
intercept=0

suspicious_samples_data =
  top_500_pca_for_sex_mislabels(qc_datasets$raw,
                                slope,
                                intercept,
                                inferred_sex_direction = 'up')
```

```{r}
pca_sex_full = suspicious_samples_data %>%
  filter(purpose == 'experiment') %>%
    ggplot(aes(PC1, PC2, 
               color = sex,
               key = subject,
               size = inferred_sex_mislabel,
               alpha = inferred_sex_mislabel)) +
      geom_point() +
      geom_abline(slope=slope,
                  intercept = intercept) +
      scale_alpha_manual(values = c('TRUE' = 1, 'FALSE' = 0.1)) +
      scale_size_manual(values = c('TRUE' = 5, 'FALSE' = 1)) +
      scale_color_manual(values = c("1" = "#B3589A",
                                    "2"="#9BBF85",
                                    "0" = "#F4A000"))

pca_sex_full

```

```{r example_pca_plot_full, include=FALSE, eval=FALSE}
#llfsRnaseq::pca_sex_qc$pca_sex_full
```


```{r pcaplot_detail}

pca_sex_suspicious_labelled = suspicious_samples_data %>%
  filter(inferred_sex_mislabel, sex != 0) %>%
  ggplot(aes(PC1, PC2,
             color = sex,
             label = ifelse(inferred_sex_mislabel & sex != 0, 
                            paste(as.character(subject), 
                                  as.character(visit), 
                                  sep="_"), NA))) +
  geom_point(aes(size = inferred_sex_mislabel, alpha = inferred_sex_mislabel)) +
  geom_text_repel(color='black') +
  geom_abline(slope=slope,
              intercept = intercept) +
  scale_alpha_manual(values = c('TRUE' = 1, 'FALSE' = 0.1)) +
  scale_size_manual(values = c('TRUE' = 5, 'FALSE' = 1)) +
  scale_color_manual(values = c("1" = "#B3589A",
                                "2"="#9BBF85",
                                "control" = "#F4A000"))

pca_sex_suspicious_labelled

```

```{r example_pca_plot_labelled, include=FALSE, eval=FALSE}
#llfsRnaseq::pca_sex_qc$pca_sex_suspicious_labelled
```

# Update the Database

**tl;dr**: review the results, update the database, and update the `dds` 
object QC as appropriate.

Ideally, these sex mislabels are already captured in the database. Here, 
we check that these samples are all labelled "suspicious_sex", that there are 
no "suspicious_sex" samples that aren't here (eg, maybe we were able to 
positively identify a sample and forgot to remove the suspicious_sex label), 
and update the database if necessary.  

Editing the database with this information can be a challenge, and requires 
that you understand the table structure. Please read the documentation 
on the database, and get comfortable with the key relationships between 
tables, and the llfs_rnaseq_metadata view.  

Except for new samples which have not been reviewed, all old samples which 
display unusual grouping should be accounted for in the `suspicious_sex` 
column of `sample` (this is `TRUE` when WGS data is unavailable), in `mislabel` 
(this is `TRUE` when the wgs data says the sample isn't what it says it is, 
but there is no positive match), or in a `sample` table `notes`, which might 
state that the sample was deemed to likely be correctly labelled for some 
other reason.  

If you notice something wonky here, it takes a good deal of investigation -- 
possibly you just need to update new sample info, most likely the 
`suspicious_sex` boolean value -- but you might also need to do some 
investigating to make sure you agree with how samples are being labelled and 
re-labelled.

## sex 1 suspicious samples
```{r}
sex_1_suspicious_samples = suspicious_samples_data %>%
  filter(inferred_sex_mislabel, inferred_sex == 1, sex !=0) %>%
  select(library_id, subject,visit,batch,sex,
         inferred_sex,suspicious_sex, mislabel,relabel,notes)
```

```{r example_sex1_mislabels, include=FALSE, eval=FALSE}
#llfsRnaseq::pca_sex_qc$sex_1_suspicious_samples
sex_1_suspicious_samples

sex_1_suspicious_samples %>% 
  left_join(llfsRnaseq::pca_sex_qc$sex_1_suspicious_samples %>% 
              select(sample_id) %>% 
              dplyr::rename(library_id=sample_id) %>% 
              mutate(prev_ident = TRUE, library_id=paste0('library_', library_id))) %>% 
  View()
```


## sex 2 suspicious samples
```{r}
sex_2_suspicious_samples = suspicious_samples_data %>%
  filter(inferred_sex_mislabel, inferred_sex == 2, sex !=0) %>%
  select(library_id, subject,visit,batch,
         sex,inferred_sex,suspicious_sex,mislabel,
         relabel, notes)
```

```{r example_sex2_mislabels, include=FALSE, eval=FALSE}
sex_2_suspicious_samples %>% 
  left_join(llfsRnaseq::pca_sex_qc$sex_2_suspicious_samples %>% 
              select(sample_id) %>% 
              dplyr::rename(library_id=sample_id) %>% 
              mutate(prev_ident = TRUE,
                     library_id=paste0('library_',
                                       library_id))) %>% 
  View()
```


```{r, include=FALSE, echo=FALSE}
# update the package data
# pca_sex_qc = list(
#   pca_sex_full = pca_sex_full,
#   pca_sex_suspicious_labelled = pca_sex_suspicious_labelled,
#   sex_1_suspicious_samples = sex_1_suspicious_samples,
#   sex_2_suspicious_samples = sex_2_suspicious_samples)
# usethis::use_data(pca_sex_qc,overwrite = TRUE)
```

